\section{Dynamic Programming from CSES}

\Topic{How to design the state}
DP is useful when choices interact, but the future only needs a compressed summary of the past. A good test: can you compute the answer for the current step using only the state variables, without looking back at individual past choices?

\begin{itemize}
	\item \textbf{Prefix DP:} sums, numbers, rows, columns. (Dice Combinations)
	\item \textbf{Boundary DP:} remember the last value/character/cell. (Array Description)
	\item \textbf{Interval DP:} process a subarray by increasing length. (Removal Game)
	\item \textbf{Subset/profile DP:} bitmask over a small set. (Elevator Rides)
	\item \textbf{Digit DP:} build a number under an upper bound. (Counting Numbers)
\end{itemize}

\Topic{Dice Combinations}

\Problem Find ordered ways to sum to $n$ using dice throws (1 to 6). Example: $n=3 \implies 4$.

\State $dp[x]$ = number of ordered dice sequences with sum $x$.
\Trans Choose the last die roll $d\in[1,6]$: add $dp[x-d]$.
\begin{lstlisting}
dp[0]=1
for x=1..n:
    for d=1..6:
        if x>=d: dp[x]+=dp[x-d]
\end{lstlisting}

\Solution Sum all valid ways to reach the previous state $x-d$. $O(n)$ time, $O(n)$ space.

\Analysis This is a basic 1D prefix DP, identical to counting compositions of $n$ into parts from $\{1..6\}$.

\Topic{Minimizing Coins}

\Problem Min coins of given denominations to make sum $x$. Example: $\{1,5,7\}, 11 \implies 3$.

\State $dp[s]$ = minimum coins needed for sum $s$.
\Trans Last coin is $c$, so previous sum is $s-c$.
\begin{lstlisting}
dp[0]=0, dp[1..x]=INF
for s=1..x:
    for c in coins:
        if s>=c:
            dp[s]=min(dp[s], dp[s-c]+1)
\end{lstlisting}

\Solution Try every coin as the last coin used. Take the minimum over valid coins. $O(nx)$ time.

\Analysis A greedy approach (largest coin) fails generally. This DP is equivalent to BFS on an implicit graph of sums.

\Topic{Coin Combinations I vs II}

\Problem \textbf{I:} Count ordered ways to make sum $x$. \textbf{II:} Count unordered ways (multisets).

\State $dp[s]$ counts ways to make sum $s$.
\begin{lstlisting}
// I: ordered
dp[0]=1
for s=1..x:
    for c in coins:
        if s>=c: dp[s]+=dp[s-c]

// II: unordered
dp[0]=1
for c in coins:
    for s=c..x:
        dp[s]+=dp[s-c]
\end{lstlisting}

\Solution In I (ordered), outer loop is sums, deciding the last coin per sequence. In II (unordered), outer loop is coins, ensuring each multiset is generated in a fixed coin order. $O(nx)$ time.

\Analysis Loop order determines permutation vs. combination counting. I counts compositions; II counts partitions.

\Topic{Removing Digits}

\Problem Find min steps to reduce $n$ to $0$ by subtracting one of its nonzero digits.

\State $dp[x]$ = minimum moves to reduce $x$ to $0$.
\Trans Pick any nonzero digit $d$ of $x$, move to $x-d$.
\begin{lstlisting}
dp[0]=0
for x=1..n:
    dp[x]=INF
    for d in digits(x):
        if d>0: dp[x]=min(dp[x],dp[x-d]+1)
\end{lstlisting}

\Solution For each number $x$, extract its digits and subtract each nonzero digit to transition from $x-d$. $O(n \cdot 10)$ time.

\Analysis Transitions go to strictly smaller numbers, so bottom-up works naturally.

\Topic{Grid Paths I}

\Problem Count paths from top-left to bottom-right of an $n \times n$ grid moving right/down, avoiding obstacles.

\State $dp[i][j]$ = number of ways to reach cell $(i,j)$.
\Trans Final move came from up or left, unless blocked.
\begin{lstlisting}
dp[0][0]=1 if open
for i=0..n-1:
    for j=0..n-1:
        if blocked: dp[i][j]=0
        else:
            if i: dp[i][j]+=dp[i-1][j]
            if j: dp[i][j]+=dp[i][j-1]
\end{lstlisting}

\Solution Paths to $(i,j)$ sum from $(i-1,j)$ and $(i,j-1)$. Blocked cells have 0 paths. $O(n^2)$ time.

\Analysis Basic 2D grid DP counting lattice paths around obstacles.

\Topic{Book Shop}

\Problem Knapsack: maximize pages from $n$ books within budget $x$. Max 1 copy per book.

\State $dp[m]$ = max pages using processed books and money $m$.
\Trans Take or skip each book.
\begin{lstlisting}
dp[0..x]=0
for book (price,pages):
    for m=x down to price:
        dp[m]=max(dp[m],dp[m-price]+pages)
\end{lstlisting}

\Solution 0/1 knapsack. Try including each book. The descending money loop ensures each book is used at most once. $O(nx)$ time.

\Analysis Ascending loop allows unbounded copies. Descending guarantees 0/1 usage by reading from previous states.

\Topic{Array Description}

\Problem Count arrays where adjacent elements differ by at most $1$. Some elements are fixed.

\State $dp[i][v]$ = ways for prefix $1..i$ ending with value $v$.
\Trans Previous value can be $v-1,v,v+1$.
\begin{lstlisting}
init i=1 from fixed/free a[1]
for i=2..n:
    for v=1..m:
        if a[i] fixed and a[i]!=v: continue
        dp[i][v]=dp[i-1][v-1]+dp[i-1][v]+dp[i-1][v+1]
\end{lstlisting}

\Solution A prefix is fully summarized by its last value. Sum transitions from adjacent allowed values. $O(nm)$ time.

\Analysis Classic boundary DP pattern since constraints are only on adjacent pairs.

\Topic{Counting Towers}

\Problem Count $2 \times n$ grid tilings with $1 \times 2$ and $2 \times 1$ blocks extending upward.

\State Small automaton: joined top block ($a$) or separated top blocks ($b$).
\Trans Extend the tower one row upward.
\begin{lstlisting}
a[1]=1, b[1]=1
for h=2..MAX:
    a[h]=2*a[h-1]+b[h-1]
    b[h]=a[h-1]+4*b[h-1]
answer(n)=a[n]+b[n]
\end{lstlisting}

\Solution A finite state machine based on the top boundary shape. The coefficients encode valid extensions. $O(N_{max})$ time.

\Analysis Classify only the boundary shape that affects the next row.

\Topic{Edit Distance}

\Problem Min inserts, deletes, or replaces to change string $a$ into $b$.

\State $dp[i][j]$ = min ops for first $i$ of $a$ into first $j$ of $b$.
\Trans Delete, insert, replace/match.
\begin{lstlisting}
dp[i][0]=i, dp[0][j]=j
for i=1..n:
    for j=1..m:
        cost = (a[i-1]!=b[j-1])
        dp[i][j]=min(dp[i-1][j]+1,
                     dp[i][j-1]+1,
                     dp[i-1][j-1]+cost)
\end{lstlisting}

\Solution Three branches for the edit choices. Match is free, mismatch costs 1. $O(nm)$ time.

\Analysis Standard 2D sequence alignment DP (Levenshtein distance).

\Topic{Rectangle Cutting}

\Problem Min cuts to split $a \times b$ rectangle into squares.

\State $dp[a][b]$ = minimum cuts to split $a\times b$.
\Trans Try every vertical and horizontal first cut.
\begin{lstlisting}
for a=1..A:
  for b=1..B:
    if a==b: dp[a][b]=0
    else:
      dp[a][b]=INF
      for x=1..a-1: dp[a][b]=min(dp[a][b],1+dp[x][b]+dp[a-x][b])
      for y=1..b-1: dp[a][b]=min(dp[a][b],1+dp[a][y]+dp[a][b-y])
\end{lstlisting}

\Solution Try all splitting lines and recursively solve the pieces. $O(ab(a+b))$ time.

\Analysis 2D interval-like DP, analogous to optimal matrix chain multiplication.

\Topic{Money Sums}

\Problem Find all achievable subset sums from $n$ coins.

\State $possible[s]$ = is sum $s$ achievable?
\Trans Take it or skip it.
\begin{lstlisting}
possible[0]=true
for coin in coins:
    for s=MAX down to coin:
        possible[s] |= possible[s-coin]
\end{lstlisting}

\Solution 0/1 knapsack decision DP. Descending loop ensures at most 1 usage per coin. $O(nS)$ time.

\Analysis Pseudo-polynomial subset sum. Bitset optimizes this beautifully.

\Topic{Removal Game}

\Problem 2-player game removing ends of an array. Maximize score difference.

\State $dp[l][r]$ = max score diff current player can force on $l..r$.
\Trans Take left or right, opponent plays remainder.
\begin{lstlisting}
dp[i][i]=a[i]
for len=2..n:
    for l=0..n-len:
        r=l+len-1
        dp[l][r]=max(a[l]-dp[l+1][r],
                     a[r]-dp[l][r-1])
answer=(sum+dp[0][n-1])/2
\end{lstlisting}

\Solution Interval DP. The difference achieved is the taken item minus the opponent's best response on the remainder. $O(n^2)$ time.

\Analysis The ``difference game'' trick reduces two scores to one difference state.

\Topic{Two Sets II}

\Problem Count ways to partition $1..n$ into two equal-sum subsets.

\State Count subsets with sum $S=n(n+1)/4$.
\Trans 0/1 subset sum over numbers $1..n$.
\begin{lstlisting}
if n*(n+1)/2 odd: answer=0
target=sum/2
dp[0]=1
for x=1..n:
    for s=target down to x:
        dp[s]+=dp[s-x]
answer=dp[target] * inv2
\end{lstlisting}

\Solution 0/1 knapsack to target sum. Divide by 2 (modular inverse) since each partition is generated twice. $O(n^3)$ time.

\Analysis Simple subset sum with a symmetry argument.

\Topic{Projects}

\Problem Weighted interval scheduling: maximize reward from non-overlapping intervals.

\State $dp[i]$ = max reward using first $i$ projects sorted by end time.
\Trans Skip $i$, or take $i$ plus best compatible prefix.
\begin{lstlisting}
sort by end
for i=1..n:
    p = last project with end[p] < start[i]
    dp[i]=max(dp[i-1], dp[p]+reward[i])
\end{lstlisting}

\Solution Binary search the last project ending before current project starts. $O(n \log n)$ time.

\Analysis Requires DP since high-weight intervals break greedy scheduling.

\Topic{Elevator Rides}

\Problem Min rides of capacity $x$ to transport $n$ people ($n \le 20$).

\State $dp[mask]=(rides,lastWeight)$ = best packing for selected mask.
\Trans Remove one person and add to previous best.
\begin{lstlisting}
dp[0]=(1,0)
for mask=1..(1<<n)-1:
    dp[mask]=(INF,INF)
    for i in mask:
        cand=dp[mask without i]
        if cand.weight+w[i] <= X: cand.weight += w[i]
        else: cand.rides++, cand.weight=w[i]
        dp[mask]=min_lex(dp[mask],cand)
\end{lstlisting}

\Solution Subset DP. Lexicographically minimizing `(rides, lastWeight)` leaves max room in the current ride. $O(n 2^n)$ time.

\Analysis Keeping the last ride as empty as possible is a greedy-optimal substructure within the DP mask.

\Topic{Counting Tilings}

\Problem Count ways to tile $n \times m$ grid with $1 \times 2$ and $2 \times 1$ dominoes.

\State $dp[col][mask]$ = ways to reach column `col` with `mask` protruding.
\Trans Fill current column by DFS.
\begin{lstlisting}
gen(row, mask, nextMask):
    if row==n: add dp[col][mask] to dp[col+1][nextMask]
    else if mask has row: gen(row+1,mask,nextMask)
    else:
        place horizontal -> nextMask has row
        if row+1 free: place vertical in current column
\end{lstlisting}

\Solution Profile DP. DFS generates valid vertical/horizontal placements transitioning between column boundaries. $O(m 4^n)$ time.

\Analysis Standard state-compression DP for grids with one small dimension.

\Topic{Counting Numbers}

\Problem Count integers in $[a,b]$ with no adjacent equal digits.

\State $f(pos,prev,tight,started)$ building left to right.
\Trans Try allowed digits, rejecting equals after start.
\begin{lstlisting}
solve(pos, prev, tight, started):
    if pos==len: return 1
    limit = tight ? digit[pos] : 9
    ans=0
    for d=0..limit:
        if started and d==prev: continue
        ans += solve(pos+1,d,tight&&d==limit,started||d>0)
\end{lstlisting}

\Solution Digit DP with state tracking the upper bound (`tight`) and leading zeros (`started`). $O(\text{digits})$ time.

\Analysis Computes $f(b) - f(a-1)$ to find the answer for the range.

\Topic{Increasing Subsequence}

\Problem Length of longest strictly increasing subsequence.

\State $tail[len]$ = smallest ending value of a length-$len$ LIS.
\begin{lstlisting}
tail = empty
for x in a:
    it = lower_bound(tail,x)
    if it==end: push x
    else *it=x
answer=tail.size()
\end{lstlisting}

\Solution Binary search in the monotonic $tail$ array and replace the smallest larger element. $O(n \log n)$ time.

\Analysis This is patience sorting. Smaller $tail$ values preserve flexibility for future extensions.
