\section{Graph Algorithms from CSES}

\Topic{Algorithm map}
First classify the graph question, then pick the tool:
\begin{itemize}
	\item \textbf{Reachability:} DFS/BFS.
	\item \textbf{Unweighted shortest path:} BFS + parent array.
	\item \textbf{Nonnegative weighted shortest path:} Dijkstra.
	\item \textbf{Negative edges/cycles:} Bellman-Ford.
	\item \textbf{All-pairs small graph:} Floyd-Warshall.
	\item \textbf{DAG:} topological sort + DP.
	\item \textbf{Mutual reachability:} SCC (Kosaraju or Tarjan).
	\item \textbf{Online connectivity:} DSU.
	\item \textbf{Capacity/matching:} Max flow (Dinic's).
\end{itemize}

\Topic{Graph representation}
\begin{lstlisting}
vector<int> adj[N];                  // unweighted
vector<pair<int,ll>> adj[N];         // weighted
vector<tuple<int,int,ll>> edges;     // Bellman/Kruskal
\end{lstlisting}
Adjacency list is best. Adjacency matrices are only for Floyd-Warshall.

\Topic{Counting Rooms}

\Problem Count connected open components in an $n \times m$ grid.

\State Grid cells are graph nodes; open neighboring cells have edges.
\begin{lstlisting}
ans=0
for each cell:
    if open and unseen:
        ans++
        BFS/DFS mark all reachable open cells
\end{lstlisting}

\Solution Iterate over cells. When finding an unseen open cell, increment count and flood-fill to mark the whole room. $O(nm)$ time.

\Analysis Connected components on an implicit grid graph.

\Topic{Labyrinth and Message Route}

\Problem Find shortest path between two points on an unweighted grid/graph.

\State BFS distance and parent from the start.
\begin{lstlisting}
q.push(start), seen[start]=true
while q not empty:
    v=q.pop()
    for u in adj[v]:
        if not seen[u]:
            seen[u]=true
            par[u]=v
            q.push(u)
if target seen: reconstruct by parent
\end{lstlisting}

\Solution BFS explores level by level, ensuring the first time a node is reached is via the shortest path. $O(V+E)$ time.

\Analysis Parent arrays encode the shortest-path tree for path reconstruction.

\Topic{Building Roads}

\Problem Find the minimum new roads to connect all $n$ cities.

\State Components of an undirected graph.
\begin{lstlisting}
leaders=[]
for v=1..n:
    if unseen[v]:
        leaders.push(v)
        DFS(v)
print leaders.size()-1
connect leaders[0] to leaders[i]
\end{lstlisting}

\Solution Find all connected components and pick a leader for each. Chain all leaders together. $O(n+m)$ time.

\Analysis $c$ components require exactly $c-1$ edges to become fully connected.

\Topic{Building Teams}

\Problem Divide pupils into two teams with no intra-team friendships (bipartite check).

\State Color each node 1 or 2.
\begin{lstlisting}
for each component:
    color[start]=1
    BFS:
        for u in adj[v]:
            if color[u]==0: color[u]=3-color[v]
            else if color[u]==color[v]: impossible
\end{lstlisting}

\Solution BFS layers alternate colors. If an edge connects same-color nodes, it's not bipartite. $O(n+m)$ time.

\Analysis Bipartite checking fails exactly when an odd-length cycle exists.

\Topic{Round Trip}

\Problem Find a cycle of length $\ge 3$ in an undirected graph.

\State DFS parent tree in an undirected graph.
\begin{lstlisting}
dfs(v,p):
    seen[v]=true
    for u in adj[v]:
        if u==p: continue
        if seen[u]: recover cycle v..u using parents
        else par[u]=v, dfs(u,v)
\end{lstlisting}

\Solution Run DFS. If we hit a visited node that isn't the parent, we found a back-edge (cycle). $O(n+m)$ time.

\Analysis Any back-edge in a DFS tree defines a fundamental cycle.

\Topic{Monsters}

\Problem Escape a grid. You and monsters move 1 cell per second. Can you reach the boundary?

\State Monster arrival time and player arrival time.
\begin{lstlisting}
multi-source BFS from all monsters -> mdist
BFS from A:
    for neighbor u:
        nt = pdist[v]+1
        if open and nt < mdist[u]:
            visit u
    if boundary reached: escape
\end{lstlisting}

\Solution Multi-source BFS computes earliest monster arrival per cell. Player BFS only steps on cells where player arrives strictly before any monster. $O(nm)$ time.

\Analysis Precomputing monster times converts dynamic enemies into static cell constraints.

\Topic{Shortest Routes I}

\Problem Shortest path from node 1 with nonnegative edge weights.

\State $dist[v]$ shortest known distance from 1.
\begin{lstlisting}
dist[1]=0
pq.push(0,1)
while pq not empty:
    (d,v)=pop min
    if d!=dist[v]: continue
    for (u,w) in adj[v]:
        if d+w < dist[u]:
            dist[u]=d+w
            pq.push(dist[u],u)
\end{lstlisting}

\Solution Dijkstra's algorithm. Processes nodes by increasing distance using a min-heap. $O((n+m)\log n)$ time.

\Analysis The greedy choice is optimal since no negative edges can provide a cheaper route later.

\Topic{Shortest Routes II}

\Problem All-pairs shortest paths on a small graph ($n \le 500$).

\State $dist[i][j]$ shortest path between all pairs.
\begin{lstlisting}
dist[i][i]=0
for edge a-b weight w:
    dist[a][b]=dist[b][a]=min(...,w)
for k=1..n:
  for i=1..n:
    for j=1..n:
      dist[i][j]=min(dist[i][j],dist[i][k]+dist[k][j])
\end{lstlisting}

\Solution Floyd-Warshall algorithm. Incrementally allows nodes $1..k$ as intermediate stops. $O(n^3)$ time.

\Analysis The outermost loop must be $k$. State transition is based on path concatenation.

\Topic{High Score and Cycle Finding}

\Problem Max weight path, or detect negative cycles.

\State Bellman-Ford distances. High Score can be solved by negating rewards.
\begin{lstlisting}
dist[1]=0
repeat n-1:
    relax all edges
mark nodes relaxed on nth round
if a relaxed node can reach n: answer is infinite
\end{lstlisting}

\Solution Bellman-Ford computes shortest paths. A relaxation on the $n$-th pass indicates a cycle. $O(nm)$ time.

\Analysis For high score, a positive cycle matters only if it's reachable from start and can reach the end.

\Topic{Flight Discount}

\Problem Min cost path, where one edge's cost can be halved.

\State $dist[v][0/1]$ = best cost to $v$ before/after using coupon.
\begin{lstlisting}
Dijkstra over (node,used)
for edge v->u weight w:
    relax (u,used) by +w
    if used==0:
        relax (u,1) by +w/2
\end{lstlisting}

\Solution Run Dijkstra on a state-expanded graph with $2n$ nodes. $O((n+m)\log n)$ time.

\Analysis Expands state space to track resource usage.

\Topic{Course Schedule}

\Problem Topological sort of $n$ courses with prerequisite edges.

\State In-degree counts remaining prerequisites.
\begin{lstlisting}
push all indegree-0 nodes
while queue not empty:
    v=pop, order.push(v)
    for u in adj[v]:
        if --indeg[u]==0: push u
if order.size()<n: IMPOSSIBLE
\end{lstlisting}

\Solution Kahn's algorithm. Process nodes with 0 in-degree, decrementing neighbors. $O(n+m)$ time.

\Analysis If all nodes aren't processed, a cycle exists.

\Topic{Longest Flight Route}

\Problem Longest path in a DAG.

\State $dp[v]$ = max number of nodes on a path from 1 to $v$ in a DAG.
\begin{lstlisting}
toposort
dp[1]=1, others=-INF
for v in order:
    for u in adj[v]:
        if dp[v]+1 > dp[u]:
            dp[u]=dp[v]+1
            par[u]=v
\end{lstlisting}

\Solution Process nodes in topological order, relaxing edges to maximize path length. $O(n+m)$ time.

\Analysis DAG property allows DP instead of NP-hard general longest path.

\Topic{Game Routes}

\Problem Count paths in a DAG from node 1 to $n$.

\State $dp[v]$ = number of paths from 1 to $v$.
\begin{lstlisting}
toposort
dp[1]=1
for v in order:
    for u in adj[v]:
        dp[u]=(dp[u]+dp[v])%MOD
\end{lstlisting}

\Solution Process in topological order, pushing path counts forward. $O(n+m)$ time.

\Analysis Standard DAG path-counting DP.

\Topic{Investigation}

\Problem Find min cost, number of min cost routes, min flights, max flights.

\State For each city: shortest distance, number of shortest routes, min flights, max flights.
\begin{lstlisting}
if nd < dist[u]:
    dist[u]=nd; ways[u]=ways[v]; mn[u]=mn[v]+1; mx[u]=mx[v]+1
else if nd == dist[u]:
    ways[u]+=ways[v]; mn[u]=min(mn[u],mn[v]+1); mx[u]=max(mx[u],mx[v]+1)
\end{lstlisting}

\Solution Run Dijkstra, but track all 4 metrics. Merge metrics on equal distances. $O((n+m)\log n)$ time.

\Analysis Aggregating path stats over shortest-path DAG.

\Topic{Planets Queries I and II}

\Problem Navigate a functional graph. 

\begin{lstlisting}
up[0][v]=next[v]
for j=1..LOG:
    up[j][v]=up[j-1][ up[j-1][v] ]
jump(v,k):
    for bit j: if k has bit: v=up[j][v]
\end{lstlisting}

\Solution Binary lifting computes $k$-th successor in $O(\log k)$ time.

\Analysis Pure binary lifting on a functional graph.

\Topic{Planets Cycles}

\Problem Find path length until a cycle repeats in a functional graph.

\State Functional graph components consist of a directed cycle with trees feeding into it.
\begin{lstlisting}
find cycle nodes by indegree pruning or DFS states
set answer for cycle nodes = cycle length
reverse edges DFS from cycle:
    ans[child] = ans[parent] + 1
\end{lstlisting}

\Solution Identify cycles, then propagate distances outwards through reversed trees. $O(n)$ time.

\Analysis "Rho-shaped" components map perfectly to this decomposition.

\Topic{Planets and Kingdoms}

\Problem Find Strongly Connected Components (SCCs).

\State Strongly connected components.
\begin{lstlisting}
DFS original graph, push by finish time
DFS reversed graph in reverse finish order
each DFS tree is one SCC
\end{lstlisting}

\Solution Kosaraju's algorithm finds SCCs by running DFS twice (once on reversed graph). $O(n+m)$ time.

\Analysis Compresses cycles to form a DAG of SCCs.

\Topic{Giant Pizza}

\Problem 2-SAT problem for pizza toppings.

\begin{lstlisting}
(a OR b) -> (!a -> b), (!b -> a)
find SCCs
if x and !x in same SCC: impossible
assignment by reverse SCC order
\end{lstlisting}

\Solution Convert to implication graph, find SCCs. If variable and negation share an SCC, unsatisfiable. $O(n+m)$ time.

\Analysis Reverse topological order of SCCs provides a valid truth assignment.

\Topic{Road Reparation}

\Problem Minimum Spanning Tree (MST).

\begin{lstlisting}
sort edges by weight
for edge (w,a,b):
    if find(a)!=find(b):
        unite(a,b)
        cost += w
\end{lstlisting}

\Solution Kruskal's algorithm adds edges greedily avoiding cycles via DSU. $O(m \log m)$ time.

\Analysis The cut property guarantees greedy choice works.

\Topic{Road Construction}

\Problem Maintain component count and max size as edges are added.

\State DSU component count and largest component size after each edge.
\begin{lstlisting}
components=n, largest=1
for road a-b:
    if unite(a,b):
        components--
        largest=max(largest,size[root])
    print components, largest
\end{lstlisting}

\Solution DSU tracks components and sizes online. $O(m \alpha(n))$ time.

\Analysis Perfect fit for disjoint set union since edges are only added.

\Topic{Mail Delivery}

\Problem Find an Eulerian circuit.

\begin{lstlisting}
if any degree odd or disconnected: IMPOSSIBLE
Hierholzer(v):
    while unused edge from v:
        mark edge used
        Hierholzer(to)
    path.push_back(v)
\end{lstlisting}

\Solution Hierholzer's algorithm traces circuits greedily, backtracking to splice them. $O(n+m)$ time.

\Analysis Exists iff all degrees are even and edges are connected.

\Topic{Download Speed}

\Problem Find Maximum Flow from 1 to $n$.

\begin{lstlisting}
while BFS builds levels:
    reset ptr
    while pushed = DFS(s, INF):
        flow += pushed
\end{lstlisting}

\Solution Dinic's algorithm pushes flow along shortest paths. $O(n^2 m)$ time.

\Analysis Standard max-flow application.

\Topic{School Dance}

\Problem Maximum Bipartite Matching.

\begin{lstlisting}
source -> boys capacity 1
boy -> girl if pair allowed
girls -> sink capacity 1
run maxflow; flow edges boy->girl are matches
\end{lstlisting}

\Solution Add source/sink with unit capacities. Run max flow. $O(m\sqrt{n})$ time.

\Analysis Solves bipartite matching natively using Dinic's.

\Topic{Hamiltonian Flights}

\Problem Count Hamiltonian paths from 1 to $n$.

\State $dp[mask][v]$ = ways to visit mask ending at $v$.
\begin{lstlisting}
dp[1<<0][0]=1
for mask containing 0:
    for v in mask:
        for u with edge v->u and u not in mask:
            dp[mask|1<<u][u]+=dp[mask][v]
answer=dp[(1<<n)-1][n-1]
\end{lstlisting}

\Solution Held-Karp bitmask DP. $O(2^n n^2)$ time.

\Analysis Tractable because $n \le 20$.
